---
title: 'Cellar Dweller: Modeling Wine Sales and Predicting Future Outlook'
author: "Alexander Harriman"
date: "11/24/2021"
output:
  word_document:
    toc: yes
  html_document:
    code_download: yes
    number_sections: yes
    theme: darkly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Data and Packages
```{r}
suppressPackageStartupMessages(library(randtests))
suppressPackageStartupMessages(library(forecast))

wine <- read.table("wine1.txt", header = FALSE)

time <- 1:length(wine$V1)

png("wine_data.png")
plot(time, wine$V1, type = "o", ylab = "Wine Sales", xlab = "Month")
dev.off()
```

# Trends

The first step is to remove any trends from the data. Since the variability increases as time increases, a logarithmic model will be used for the trend.

```{r}
logModel <- lm(V1 ~ log(time), data = wine)
summary(logModel)

#The log(time) variable is statistically significant and will remain in the final model.

#Checking residuals:
logResid <- resid(logModel)
plot(time, logResid, type = "o")
```

The second step is to remove seasonal components from the data. Since the residuals for the logarithmic model appear to have a seasonal trend with period d = 12 (which fits the original data's seasonal trend), both it and d = 6 will be tested.

```{r}
logTime <- lm(V1 ~ log(time) + cos(2*pi*time/12) + sin(2*pi*time/12) + cos(2*pi*time/6) + sin(2*pi*time/6), data = wine)

summary(logTime)

#Since both terms with d = 6 are not statistically significant, the model will be run again without them

logTime2 <- lm(V1 ~ log(time) + cos(2*pi*time/12) + sin(2*pi*time/12), data = wine)
summary(logTime2)

#The terms with period d = 12 remain in the model.

##Residuals:
timeResid <- resid(logTime2)

png("reg_resid.png")
plot(time, timeResid, type = "o")
dev.off()
```

# Diagnostics

Now that the trend and seasonal components are removed, the first round of diagnostics will determine if the residuals are IID using the residual tests.

```{r}
#ACF and PACF Visualization Test
png("reg_acf.png")
acf(timeResid, lag.max = 50)
dev.off()
png("reg_pacf.png")
pacf(timeResid, lag.max = 50)
dev.off()

#Neither of these models have lag values within the boundaries, so both tests fail.

#Tests
Box.test(timeResid, type = "Ljung-Box")
turning.point.test(timeResid)
rank.test(timeResid)
difference.sign.test(timeResid)

#The Box Test and the Rank Test have p-values that reject the null, and the Turning Point Test and Difference Sign Test have p-values that accept the null.

#Since most of the tests reject the null, the residuals are not IID.
```

# ARMA Modeling

The next step is to find the best ARMA model. 

Based on the ACF, an MA model is not proper since values of the ACF are continuously outside of the bounds.

Based on the PACF, an AR model is a better model, but is not strong. We can use the ar() function to determine the best model:

```{r}
#Yule-Walker
yuleMod <- ar(timeResid, method = "yule-walker")
yuleMod

#MLE
mleMod <- ar(timeResid, method = "mle")
mleMod

#The MLE has an AR(12) model and the Yule-Walker has an AR(13) model.

#The best ARMA model:
arimaMod <- auto.arima(timeResid)
arimaMod

#The best ARMA model is ARIMA(0,1,1) using auto.arima. This simplifies to an MA(1) model with the I term removed.
```

# Residual Diagnostics

All of the candidate functions will now go through the second round of residual diagnostics.

## AR(13) for Yule-Walker
```{r}
yuleResid <- resid(yuleMod)

png("yule_resid.png")
plot(yuleResid, type = "o")
dev.off()

#The first 13 values need to be removed since this is an AR(13) model
yuleResid <- yuleResid[14:length(yuleResid)]

#Tests
png("yule_acf.png")
acf(yuleResid, lag.max = 50)
dev.off()
png("yule_pacf.png")
pacf(yuleResid, lag.max = 50)
dev.off()

Box.test(yuleResid, type = "Ljung-Box")
turning.point.test(yuleResid)
rank.test(yuleResid)
difference.sign.test(yuleResid)

#Both the ACF and PACF have only a few lag values that barely cross over the boundary lines. So, these tests are inconclusive.

##However, since each test but the Rank Test accepts the null, we can conclude that the residuals for the Yule-Walker estimation are IID. 
```

## AR(12) for MLE
```{r}
mleResid <- resid(mleMod)

png("mle_resid.png")
plot(mleResid, type = "o")
dev.off()

#The first 12 values need to be removed since this is an AR(12) model
mleResid <- mleResid[13:length(mleResid)]

#Tests
png("mle_acf.png")
acf(mleResid, lag.max = 50)
dev.off()
png("mle_pacf.png")
pacf(mleResid, lag.max = 50)
dev.off()

Box.test(mleResid, type = "Ljung-Box")
turning.point.test(mleResid)
rank.test(mleResid)
difference.sign.test(mleResid)

#Just like the Yule-Walker estimation, the ACF and PACF tests are inconclusive, but every test but the Rank Test accepts the null. So, the MLE estimation has residuals that are IID.
```

## ARIMA(0,1,1) for auto.arima()
```{r}
autoResid <- resid(arimaMod)

#Tests
acf(autoResid, lag.max = 50)
pacf(autoResid, lag.max = 50)

Box.test(autoResid, type = "Ljung-Box")
turning.point.test(autoResid)
rank.test(autoResid)
difference.sign.test(autoResid)

#For this model, the ACF test fails due to many of the ACF values falling outside of the bounds. The PACF is inconclusive, but to a lesser extent than both AR-only models.

##All tests except for the Rank test accept the null hypothesis. So, it is unclear whether the residuals for this ARIMA(0,1,1) model are IID or not.
```

# The Best Model
Since fewer tests support the ARIMA(0,1,1) model's residuals being IID, it drops out of the analysis. So, that leaves the two AR(p) models estimated by Yule-Walker and MLE.

In order to determine the best model, the AICC values for both models will be investigated. Although the MLE model has a smaller p value, the AICC may find the AR(13) better.

```{r}
#Verifying the Model using Arima()

yuleMA <- Arima(timeResid, order = c(13,0,0)) #Yule-Walker
yuleMA

mleMA <- Arima(timeResid, order = c(12,0,0)) #MLE
mleMA
```

# Conclusion
The AICC value for the AR(13) model is less than the AICC value for the AR(12) model. However, since the two are relatively similar, both models could be chosen. 

Overall, since 12 < 13, the best choice to model the wine data is **an AR(12) time series with a seasonal time series of period d = 12 and a logarithmic trend.**

The model is estimated to be **X_t = 337.2log(t) - 250.39cos(2pi(t)/12) - 273.82sin(2pi(t))/12) + E_t**, 
where E_t is an AR(12) model of the form **E_t = 0.1924E_(t-1) + 0.1067E_(t-2) + 0.1674E_(t-3) + 0.0739E_(t-4) + 0.0849E_(t-5) - 0.1072E_(t-6) + 0.1128E_(t-7) + 0.0616E_(t-8) - 0.0034E_(t-9) - 0.1268E_(t-10) - 0.141E_(t-11) + 0.5053E_(t-12) + Z_t**, 
where **Z_t ~ IID(0, 45876)**, 
based on an MLE estimation.


# Predicting the Next 5 Values
Using the model, we can predict future values. 

```{r}
#Vector Initialization
seasonal <- matrix(data = rep(0,5), nrow = 5, ncol = 1)
log <- matrix(data = rep(0,5), nrow = 5, ncol = 1)

#Predicting the AR(12) portion
arPred <- predict(mleMod, n.ahead = 5)
arPred #Error values

#Predicting the Seasonal Trend:
for(i in 143:147){
  seasonal[i-142] <- (-250.39*(cos(2*pi*i/12)) - 273.82*sin(2*pi*i/12))
}

#Predicting the Logarithmic Trend:
for(i in 143:147){
  log[i-142] <- (337.2*log(i))
}

#Combining the three predictions to get final values:
for(i in 1:5){
  print(arPred$pred[i] + seasonal[i] + log[i]) #X_t values
}

```

So, the next five values for wine sales are 1968.216, 2123.137, 1242.278, 1664.179, and 1761.134. These values make sense based on the original data; they fit into the general values around the end of the original data.

# Plotting the Extra Data Values
```{r}
#Adding the Estimated Values to Original Data
predictedValues <- c(1968.216, 2123.137, 1242.278, 1664.179, 1761.134)
allValues <- c(wine$V1, predictedValues)

allTimes <- 1:147

#Total Plot
png("all_data.png")
plot(allTimes, allValues, type = "o", xlab = "Time", ylab = "Wine Sales")
dev.off()
```